---
title: "Basic time series"
author: "Steve Simon"
date: "3/4/2019"
output: powerpoint_presentation
---

### Introduction

* Keep things simple
  + Evenly spaced
  + No gaps
  + Single series
* Topics covered
  + Structure
  + Plotting
  + Detrending and deseasonalizing
  + Basic Box-Jenkins models
  
### Structure

The base R package has an object class, ts. You can specify three parameters: start, stop, and frequency. 

### Structure of co2 data

```{r co2}
str(co2)
attributes(co2)
```

### Plot of co2 data

```{r co2-plot, echo=FALSE}
plot(co2)
```

### Structure of LakeHuron data

```{r LakeHuron}
str(LakeHuron)
attributes(LakeHuron)
```

### Plot of LakeHuron data

```{r LakeHuron-plot}
plot(LakeHuron)
```

### Structure of mdeaths data

```{r mdeaths}
str(mdeaths)
attributes(mdeaths)
```

### Plot of mdeaths

```{r mdeaths-plot}
plot(mdeaths)
```

### Structure of Nile data

```{r Nile}
str(Nile)
attributes(Nile)
```

### Plot of Nile data

```{r Nile-plot, echo=FALSE}
plot(Nile)
```

### Structure of nottem

```{r nottem}
str(nottem)
attributes(nottem)
```

### Plot of nottem

```{r nottem-plot, echo=FALSE}
plot(nottem)
```

### Structure of sunspot.year

```{r sunspot}
str(sunspot.year)
attributes(sunspot.year)
```

### Plot of sunspot.year

```{r sunspot-plot}
plot(sunspot.year)
```

### Structure of treering

```{r treering}
str(treering)
attributes(treering)
```

### Plot of treering

```{r treering-plot, echo=FALSE}
plot(treering)
```

### UK time series

```{r UK}
str(UKDriverDeaths)
str(UKgas)
```

### US time series

```{r US}
str(USAccDeaths)
attributes(USAccDeaths)
```

### USAccDeaths plot

```{r USAccDeaths-plot, echo=FALSE}
plot()