---
title: "Basic time series"
author: "Steve Simon"
date: "3/4/2019"
output: powerpoint_presentation
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

### Introduction

* Keep things simple
  + Evenly spaced
  + No gaps
  + Single series
* Topics covered
  + Structure
  + Plotting
  + Detrending and deseasonalizing
  + Basic Box-Jenkins models
  
### White noise

```{r white-noise}
set.seed(23571113)
plot(rnorm(100), type="l")
```

### Structure

The base R package has an object class, ts. You can specify three parameters: start, stop, and frequency. 

### UKgas

```{r UKgas, echo=TRUE}
str(UKgas)
attributes(UKgas)
```

### Plot UKGas

```{r UKgas-plot}
plot(UKgas)
```

### Structure of co2 data

```{r co2, echo=TRUE}
str(co2)
attributes(co2)
```

### Plot of co2 data

```{r co2-plot}
plot(co2)
```

### Decomposition of co2

```{r co2-decomp, echo=TRUE}
co2_d <- decompose(co2)
str(co2_d)
```

### Plot of co2 decomposition

```{r co2-decomp-plot}
plot(co2_d)
```

### UKgas decomposition

```{r UKgas-decomp, echo=TRUE}
UKgas_d <- decompose(UKgas, type="multiplicative")
str(UKgas_d)
```

### Plot of UKgas decomposition

```{r UKgas-decomp-plot}
plot(UKgas_d)
```

### Autocorrelations of co2 data

```{r co2-auto}
acf(co2)
```

### Partial autocorrelations of co2 data

```{r co2-pacf}
pacf(co2)
```

### Structure of LakeHuron data

```{r LakeHuron, echo=TRUE}
str(LakeHuron)
attributes(LakeHuron)
```

### Plot of LakeHuron data

```{r LakeHuron-plot}
plot(LakeHuron)
```

### Structure of mdeaths data

```{r mdeaths, echo=TRUE}
str(mdeaths)
attributes(mdeaths)
```

### Plot of mdeaths

```{r mdeaths-plot}
plot(mdeaths)
```

### Structure of Nile data

```{r Nile, echo=TRUE}
str(Nile)
attributes(Nile)
```

### Plot of Nile data

```{r Nile-plot}
plot(Nile)
```

### Structure of nottem

```{r nottem, echo=TRUE}
str(nottem)
attributes(nottem)
```

### Plot of nottem

```{r nottem-plot}
plot(nottem)
```

### Structure of sunspot.year

```{r sunspot, echo=TRUE}
str(sunspot.year)
attributes(sunspot.year)
```

### Plot of sunspot.year

```{r sunspot-plot}
plot(sunspot.year)
```

### Structure of treering

```{r treering, echo=TRUE}
str(treering)
attributes(treering)
```

### Plot of treering

```{r treering-plot}
plot(treering)
```

### UKDriverDeaths

```{r UK, echo=TRUE}
str(UKDriverDeaths)
attributes(UKDriverDeaths)
```

### Plot UKDriver Deaths

```{r UKDriverDeaths-plot}
plot(UKDriverDeaths)
```

### USAccDeaths

```{r US, echo=TRUE}
str(USAccDeaths)
attributes(USAccDeaths)
```

### USAccDeaths plot

```{r USAccDeaths-plot, echo=FALSE}
plot(USAccDeaths)
```