---
title: "Basic time series"
author: "Steve Simon"
date: "3/4/2019"
output: powerpoint_presentation
header-includes :
- \ usepackage { amsmath }
- \ usepackage { mathtools }
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, fig.width=9)
n_sim <- 1000
r <- 0.5
```

### Introduction

* Keep things simple
  + Evenly spaced
  + No gaps
  + Single long series
* Topics covered
  + Structure
  + Plotting
  + Detrending and deseasonalizing
  + Basic Box-Jenkins models
  
### Decomposing a time series

* Additive decomposition
  + Y - original data
  + T - time trend
  + S - seasonal component
  + Z = random component

$Y_t = T_t + S_t + Z_t$

* Multiplicative decomposition
  + (or use a log transformation)

$Y_t = T_t \times S_t \times Z_t$

### Differencing

An alternate way to remove a trend or deseasonalize a data set is through differencing.

* Detrending

$D_t = Y_t - Y_{t-1}$

* Deseasonalizing (s=# of seasons)

$D_t = Y_t - Y_{t-s}$

### Models for the random component

White noise

* $Z_t = \epsilon_t$

Autoregressive (AR)

* $Z_t = \epsilon_t + \phi Z_{t-1}$

Moving average (MA)

* $Z_t = \epsilon_t+\theta \epsilon_{t-1}$

### Autocorrelation and partial autocorrelation

* acf function
  + Plots corr($Z_t,Z_{t-k}$) for k=1, 2, ...

* pacf function
  + Plots corr($Z_t,Z_{t-k}|Z_{t-1},...,Z_{t-k+1}$) for k=2, 3, ...

### Structure

The base R package has an object class, ts. You can specify three parameters: start, stop, and frequency. You do not need to place your data in a ts object before running time series analyses. 

There is a new class, tsibble, that I have not had a chance to work with. It is based on the principles of tidy data. Read the [tsibble vignette](https://cran.rstudio.com/web/packages/tsibble/vignettes/intro-tsibble.html) for more information.

### Structure of co2 data

```{r co2, echo=TRUE}
str(co2)
attributes(co2)
```

### Plot of co2 data

```{r co2-plot}
plot(co2)
```

### Decomposition of co2

```{r co2-decomp, echo=TRUE}
co2_d <- decompose(co2)
str(co2_d)
```

### Plot of co2 decomposition

```{r co2-decomp-plot}
plot(co2_d)
```

### Deseasonalizing co2

```{r co2-deseasonal}
plot(diff(co2, lag=12))
```

### Detrending co2

```{r co2-detrend}
plot(diff(co2, lag=1))
```

### Both Detrending  and deseasonalizing co2

```{r co2-both}
plot(diff(diff(co2, lag=12), lag=1))
```


### UKgas

```{r UKgas, echo=TRUE}
str(UKgas)
attributes(UKgas)
```

### Plot UKgas

```{r UKgas-plot}
plot(UKgas)
```

### Plot UKgas on a log scale

```{r UKgas-log}
plot(log(UKgas))
```

### UKgas decomposition

```{r UKgas-decomp, echo=TRUE}
UKgas_d <- decompose(log(UKgas))
str(UKgas_d)
```

### Plot of UKgas decomposition

```{r UKgas-decomp-plot}
plot(UKgas_d)
```

### Plot ofUKgas difference model

```{r UKgas-diff}
plot(diff(diff(log(UKgas), lag=4), lag=1))
```

### White noise

* $Z_t = \epsilon_t$
* Ideal pattern
  + Nothing left to predict
  + Any deviation from white noise is an opportunity

### A simulation of white noise

```{r white-noise}
set.seed(23571113)
white_noise <- arima.sim(n=n_sim, list())
plot(white_noise, type="l")
```

### Theoretical correlation matrix for white noise

```{r identity}
m <- diag(6)
m
```

### Autoregressive model

ARIMA(1, 0, 0)

* $Z_t=\phi Z_{t-1} + \epsilon_t$

The autoregressive model produces a "ripple effect" because this model implies a relationship going back in time.

* $Z_t=\phi^2 Z_{t-2}+\phi \epsilon_{t-1} + \epsilon_t$

and $Z_{t-3}$

* $Z_t=\phi^3 Z_{t-3}  + \phi^2 \epsilon_{t-2} + \phi \epsilon_{t-1.000} + \epsilon_t$

and so on.

### Theoretical correlation matrix for positive autoregessive model

```{r pos-ar-corr}
m <- matrix(0, nrow=6, ncol=6)
for (i in 1:6) {for (j in 1:6) {m[i,j] <- r**abs(i-j)}}
round(m, 3)
```

### A simulation of a positive autoregressive model

```{r pos-ar-sim}
set.seed(23571113)
pos_ar_sim <- arima.sim(n=n_sim, list(ar=r))
par(mfrow=c(2, 1), mar=c(0, 0, 2.6, 0))
plot(pos_ar_sim, main="Positive autoregressive time series", axes=FALSE)
plot(white_noise, main="White noise", axes=FALSE)
```

### Auotcorrelation patterns for positive autoregressive model

```{r pos-ar-acf}
par(mfrow=c(1, 2))
acf(pos_ar_sim, na.action=na.omit, main="acf(pos_ar_sim)")
pacf(pos_ar_sim, na.action=na.omit, main="pacf(pos_ar_sim)")
```

### Theoretical correlation matrix for negative autoregessive model

```{r neg-ar-corr}
m <- matrix(0, nrow=6, ncol=6)
for (i in 1:6) {for (j in 1:6) {m[i,j] <- (-r)**abs(i-j)}}
round(m, 3)
```

### A simulation of a negative autoregressive model

```{r neg-ar-sim}
set.seed(23571113)
neg_ar_sim <- arima.sim(n=n_sim, list(ar=-r))
par(mfrow=c(2, 1), mar=c(0, 0, 2.6, 0))
plot(neg_ar_sim, main="Negative autoregressive time series", axes=FALSE)
plot(white_noise, main="White noise", axes=FALSE)
```

### Auotcorrelation patterns for negative autoregressive model

```{r neg-ar-acf}
par(mfrow=c(1, 2))
acf(neg_ar_sim, na.action=na.omit, main="acf(neg_ar_sim)")
pacf(neg_ar_sim, na.action=na.omit, main="pacf(neg_ar_sim)")
```

### Moving average model

ARIMA(0, 0, 1)

* $Z_t = \theta \epsilon_{t-1} + \epsilon_t$

There is no "ripple effect" with a moving average model.

### Theoretical correlation matrix for positive moving average model

```{r pos-ma-corr}
m <- diag(6)
for (i in 1:5) {
  m[i, i+1] <- r
  m[i+1, i] <- r
}
round(m, 1)
```

### A simulation of a positive moving average model

```{r pos-ma-sim}
set.seed(23571113)
pos_ma_sim <- arima.sim(n=n_sim, list(ma=r))
par(mfrow=c(2, 1), mar=c(0, 0, 2.6, 0))
plot(pos_ma_sim, main="Positive moving average time series", axes=FALSE)
plot(white_noise, main="White noise", axes=FALSE)
```

### Auotcorrelation patterns for positive moving average model

```{r pos-ma-acf}
par(mfrow=c(1, 2))
acf(pos_ma_sim, na.action=na.omit, main="acf(pos_ma_sim)")
pacf(pos_ma_sim, na.action=na.omit, main="pacf(pos_ma_sim)")
```

### Theoretical correlation matrix for negative moving average model

```{r neg-ma-corr}
m <- diag(6)
for (i in 1:5) {
  m[i, i+1] <- -r
  m[i+1, i] <- -r
}
round(m, 1)
```

### A simulation of a negative moving average model

```{r neg-ma-sim}
set.seed(23571113)
neg_ma_sim <- arima.sim(n=n_sim, list(ma=-r))
par(mfrow=c(2, 1), mar=c(0, 0, 2.6, 0))
plot(neg_ma_sim, main="Negative moving average time series", axes=FALSE)
plot(white_noise, main="White noise", axes=FALSE)

```

### Auotcorrelation patterns for positive moving average model

```{r neg-ma-acf}
par(mfrow=c(1, 2))
acf(neg_ma_sim, na.action=na.omit, main="acf(neg_ma_sim)")
pacf(neg_ma_sim, na.action=na.omit, main="pacf(neg_ma_sim)")
```

### More complex autoregressive patterns

ARIMA(2, 0, 0)

* $Z_t = \phi_1 Z_{t-1} + \phi_2 Z_{t-2} + \epsilon_t$

ARIMA(0, 0, 2)

* $Z_t = \theta_2 \epsilon_{t-2} + \theta_1 \epsilon_{t-1} + \epsilon_t$

ARMA(1, 0, 1)

* $Z_t = \phi_1 Z_{t-1} + \theta_1 \epsilon_{t-1} + \epsilon_t$

ARIMA(1, 1, 1)

* $D_t = \phi_1 D_{t-1} + \theta_1 \epsilon_{t-1} + \epsilon_t$


### acf and pacf for co2 data

```{r co2-auto}
par(mfrow=c(1, 2))
acf(co2_d$random, na.action=na.omit, main="acf(co2_d$random)")
pacf(co2_d$random, na.action=na.omit, main="pacf(co2_d$random)")
```

### Structure of LakeHuron data

```{r LakeHuron, echo=TRUE}
str(LakeHuron)
attributes(LakeHuron)
```

* Note: Annual data has no seasonality.

### Plot of LakeHuron data

```{r LakeHuron-plot}
plot(LakeHuron)
```

### acf and pacf for data

```{r LakeHuron-auto}
par(mfrow=c(1, 2))
acf(LakeHuron, na.action=na.omit)
pacf(LakeHuron, na.action=na.omit)
```

### Plot of LakeHuron data

```{r diff-LakeHuron-plot}
plot(diff(LakeHuron))
```

### acf and pacf for data

```{r diff-LakeHuron-auto}
par(mfrow=c(1, 2))
acf(diff(LakeHuron), na.action=na.omit)
pacf(diff(LakeHuron), na.action=na.omit)
```

### Structure of mdeaths data

```{r mdeaths, echo=TRUE}
str(mdeaths)
attributes(mdeaths)
```

### Plot of mdeaths

```{r mdeaths-plot}
plot(mdeaths)
```

### Detrend and deseasonalize mdeaths

```{r mdeaths-decomp}
mdeaths_d <- decompose(mdeaths)
plot(mdeaths_d)
```

### acf and pacf for mdeaths data

```{r mdeaths-auto}
par(mfrow=c(1, 2))
acf(mdeaths_d$random, na.action=na.omit, main="acf(mdeaths_d$random)")
pacf(mdeaths_d$random, na.action=na.omit, main="pacf(mdeaths_d$random)")
```

### Structure of Nile data

```{r Nile, echo=TRUE}
str(Nile)
attributes(Nile)
```

### Plot of Nile data

```{r Nile-plot}
plot(Nile)
```

### acf and pacf for Nile data

```{r Nile-auto}
par(mfrow=c(1, 2))
acf(Nile, na.action=na.omit, main="acf(Nile)")
pacf(Nile, na.action=na.omit, main="pacf(Nile)")
```

### Structure of nottem

```{r nottem, echo=TRUE}
str(nottem)
attributes(nottem)
```

### Plot of nottem

```{r nottem-plot}
plot(nottem)
```

### Detrend and deseasonalize nottem

```{r nottem-decomp}
nottem_d <- decompose(nottem)
plot(nottem_d)
```

### acf and pacf for nottem data

```{r nottem-auto}
par(mfrow=c(1, 2))
acf(nottem_d$random, na.action=na.omit, main="acf(nottem_d$random)")
pacf(nottem_d$random, na.action=na.omit, main="pacf(nottem_d$random)")
```

### Structure of sunspot.year

```{r sunspot, echo=TRUE}
str(sunspot.year)
attributes(sunspot.year)
```

### Plot of sunspot.year

```{r sunspot-plot}
plot(sunspot.year)
```

### Structure of treering

```{r treering, echo=TRUE}
str(treering)
attributes(treering)
```

### Plot of treering

```{r treering-plot}
plot(treering)
```

### acf and pacf for treering data

```{r treering-auto}
par(mfrow=c(1, 2))
acf(treering, na.action=na.omit, main="acf(treering)")
pacf(treering, na.action=na.omit, main="pacf(treering)")
```

### acf and pacf for treering data

```{r diff-treering-auto}
par(mfrow=c(1, 2))
acf(diff(treering), na.action=na.omit, main="acf(diff(treering))")
pacf(diff(treering), na.action=na.omit, main="pacf(diff(treering))")
```

### UKDriverDeaths

```{r UK, echo=TRUE}
str(UKDriverDeaths)
attributes(UKDriverDeaths)
```

### Plot UKDriverDeaths

```{r UKDriverDeaths-plot}
plot(UKDriverDeaths)
```

### Detrend and deseasonalize UKDriverDeaths

```{r UKDriverDeaths-decomp}
UKDriverDeaths_d <- decompose(UKDriverDeaths)
plot(UKDriverDeaths_d)
```

### acf and pacf for UKDriverDeaths data

```{r UKDriverDeaths-auto}
par(mfrow=c(1, 2))
acf(UKDriverDeaths_d$random, na.action=na.omit, main="acf(UKDriverDeaths_d$random)")
pacf(UKDriverDeaths_d$random, na.action=na.omit, main="pacf(UKDriverDeaths_d$random)")
```

### USAccDeaths

```{r US, echo=TRUE}
str(USAccDeaths)
attributes(USAccDeaths)
```

### USAccDeaths plot

```{r USAccDeaths-plot, echo=FALSE}
plot(USAccDeaths)
```

### Detrend and deseasonalize USAccDeaths

```{r USAccDeaths-decomp}
USAccDeaths_d <- decompose(USAccDeaths)
plot(USAccDeaths_d)
```

### acf and pacf for USAccDeaths data

```{r USAccDeaths-auto}
par(mfrow=c(1, 2))
acf(USAccDeaths_d$random, na.action=na.omit, main="acf(USAccDeaths_d$random)")
pacf(USAccDeaths_d$random, na.action=na.omit, main="pacf(USAccDeaths_d$random)")
```

